Coastal Flooding (surface water, highway, sewer and watercourse flooding) across Torbay, UK

### 2017-08-15 EU-Circle/eu-circle_case-study-3.Rmd
EU-Circle case study 3 - updated analysis

@@

@@

**Wind**

Synoptic storms are associated with winds, storm surges and waves.

Plot the number of trajectories over time

```{r}
plot(cyclones.uk,new=FALSE)
```

---------------------------------

@@

Load the data and weed out stations with lots of missing values

@@

Fill in missing values, but since there are too many missing values, we need to get some more data from the region. This is ok since we are working with annual aggregated values, which may imply larer spatial scales. In any case, these are used to assist the interpolation for the MIDAS sations.

@@

supplement these to the MIDAS data:

@@

```{r}
plot(fw); grid()
for (i in 1:dim(mu)[2]) lines(trend(mu[,i]))
```

```{r probabilities, warning=FALSE, message=FALSE}
Pr.gt.30 <- PrexpPr(mu,fw,x0=30)
plot(Pr.gt.30); grid()
for (i in 1:dim(mu)[2]) lines(trend(Pr.gt.30[,i]))
```

@@

Another attempt to search for systematic influences on the precipitation statistics ($\mu$) is to examine the mean seasonal cycle. This type of approach may

@@

```{r precipitation-intensity, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the wet-day mean precipitation
plot(MU)
grid()
```

There is some indication of a mean seasonal cycle in $\mu$, with minimum values in late winter spring and higher values in late summer and autumn.

```{r wet-day frequency, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the wet-day frequency
plot(FW)
grid()
```

There is a clear seasonal cycle in the wet-day occurrence - more rainy days in the uatumn.

```{r sesonal-precipitation, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the total precipitation
MP <- 30*MP
attr(MP,'unit') <- 'mm/month'
plot(MP)
grid()
```

@@

```{r winter-rain, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Estimate the PCA for the annual cycle to identify the general features:
pca.MU <- PCA(MU,n=10)
plot(pca.MU)
```

Most stations have the same features when it comes to a mean seasonal cycle in $\mu$, with the leading PCA accounting for 80% of the variance.

@@

## EOBS gridded data

@@

```{r predictand, warning=FALSE, message=FALSE}
pr.eobs <- retrieve('data.ECAD/rr_0.25deg_reg.nc',lon=c(-6,-1),lat=c(50,52))
mu <- annual(pr.eobs,FUN='wetmean')
fw <- annual(pr.eobs,FUN='wetfreq')
```

@@

```{r trend, warning=FALSE, message=FALSE}
## The area mean wet-day mean precipitation
plot(mu,new=FALSE)
## The area mean wet-day frequency
plot(fw,new=FALSE)
## The early part of the record looks suspect:
mu <- subset(mu,it=c(1958,2015))
fw <- subset(fw,it=c(1958,2015))
## Show maps of trends for the wet-day mean precipitation
map(mu,FUN='trend',
    colbar=list(breaks=seq(-0.12,0.12,by=0.02),pal='t2m',rev=TRUE))
## Map of trends in wet-day frequency:
map(fw,FUN='trend',
    colbar=list(breaks=seq(-0.01,0.01,by=0.001),pal='t2m',rev=TRUE))
```

@@

```{r EOFs-in-precip-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The area mean wet-day mean precipitation
eof.mu <- EOF(mu)
## The area mean wet-day frequency
plot(eof.mu)

@@

```{r EOFs-in-precip-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Show maps of trends for the wet-day mean precipitation
eof.fw <- EOF(fw)
## Map of trends in wet-day frequency:
plot(eof.fw)
```

@@

@@

```{r DS, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The wet-day frequency and SLP
ds.fw <- DS(subset(eof.fw,ip=1:4),eof.slp,eofs=1:5)
plot(ds.fw,new=FALSE)
ds.mu <- DS(subset(eof.mu,ip=1:4),eof.es,eofs=1:7)
plot(ds.mu,new=FALSE)
```

@@

```{r residual-for-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The residual of the wet-day precipitation
z.mu <- as.field(ds.mu)
#res.mu <- as.residual(z.mu)
res.mu <- mu - z.mu
res.mu <- attrcp(mu,res.mu)
class(res.mu) <- class(mu)
eof.res.mu <- EOF(res.mu)
plot(eof.res.mu)
```

@@

For comparison the original data look like this:
```{r EOF-for-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The original of the wet-day precipitation
plot(EOF(mu))
```

```{r residual-for-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Repeat with the wet-day frequency for the benefit of comparison
z.fw <- as.field(ds.fw)
#res.fw <- as.residual(ds.fw)
## Another way to estimate the residuals:
res.fw <- fw - z.fw
res.fw <- attrcp(fw,res.fw)
class(res.fw) <- class(fw)
eof.res.fw <- EOF(res.fw)
plot(eof.res.fw)
```

@@

For comparison the original data look like this:
```{r EOF-for-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The original of the wet-day precipitation
plot(EOF(fw))
```

@@

Analyse the sea-level trends and connection to global temperatures to assess future outcome. First the stations from the GLOSS dataset (global monthly data):

@@

```{r}
plot(gloss)
grid()
```

@@

```{r}
plot(newlyn)
grid()
```

@@

```{r sonel, warning=FALSE, message=FALSE}
sonel <- aggregate(station.sonel(),year,FUN='max',na.rm=TRUE)
plot(sonel)
grid()
```

@@

```{r combine-sea-level, warning=FALSE, message=FALSE}
index(newlyn) <- year(newlyn)
index(sonel) <- year(sonel)
index(gloss) <- year(gloss)
Z <- combine.stations(newlyn,sonel,gloss)
nv <- apply(coredata(Z),2,FUN='nv')
Z <- subset(Z,is=(nv > 50))
Z <- subset(Z,it=c(1966,2015))
coredata(Z)[!is.finite(Z)] <- NA
coredata(Z) <- t(t(coredata(Z)) - colMeans(coredata(Z),na.rm=TRUE))
plot(Z); grid()
```

@@

```{r sl-trend-map, warning=FALSE, message=FALSE, fig.height=8}
zc <- coredata(Z)
zc[is.infinite(zc)] <- NA
zc -> coredata(Z)
map(Z,pch=19,FUN='trend',cex=-2,add.text=TRUE)
```

@@

@@

@@

```{r slp-trend-map, warning=FALSE, message=FALSE}
slp.eobs <- retrieve('data.ECAD/pp_0.25deg_reg.nc',lon=c(-6,0),lat=c(48,52))
slp.eobs <- 10000*slp.eobs ## The data have some funny scaling factor
attr(slp.eobs,'unit') <- 'Pa'
map(slp.eobs,FUN='trend')
```

@@

```{r trend-map, warning=FALSE, message=FALSE}
## Fudge it - pick some windstations, but do not use their data. Reset the
## coordinates to select a triangle of SLP from the EOBS data.
ss <- select.station(param='slp',src='ecad',cntr=c('France','UNITED KINGDOM'))
fx <- station(ss)
triangle <- subset(fx,is=1:3)
## The stations are not situated quite where there is valid slp data in the EOBS dataset
## It's only their coordinates that is used for interpolating SLP for three stations
## Using station objects as is-object makes the results a station object
attr(triangle,'longitude') <- c(-4,-2,-1.25)
attr(triangle,'latitude') <- c(50.50,51,49)
pp.eobs <- regrid(slp.eobs,is=triangle)
## Use triagulation for stations to estimate the wind
uv <- geostrophicwind(pp.eobs)
plot(annual(uv,nmin=300),main='zonal and meridional wind components',
     sub='Estimated from triangulation of mean sea-level pressure from EOBS',zoom=7)
grid()
```

The estimated annual mean geostrophic wind reveals that something changed between 2006 and 2010. The change seems to have taken place over 4 years.

```{r trend-event-2006, warning=FALSE, message=FALSE}
plot(annual(uv,nmin=300),main='zonal and meridional wind components',
     xlim=c(2000,2010),map.show=FALSE)
grid()
```

This event also is seen in the upper tail of the wind speed distribution. There are four features concerning these wind speed estimates: a downward trend from moderate high wind speeds in the 1950s, two years with a large drop in values in the 1970s, a number of years with high values around year 2010, and a slight increasing trend in the upper wind speeds over the record.

```{r wind-speed-trend, warning=FALSE, message=FALSE}
## Trend in upper wind speed
ff <- sqrt(uv[,1]^2 + uv[,2]^2)
plot(annual(ff,"quantile",probs=0.98),lwd=2,
     main='Wind speed: annual 98-percentile (m/s)')
lines(trend(annual(ff,"quantile",probs=0.98)),lty=2)
grid()
trend.coef(annual(ff,"quantile",probs=0.98))
trend.err(annual(ff,"quantile",probs=0.98))
```

@@

```{r wind-direction-trend, warning=FALSE, message=FALSE}
## Trend in upper wind direction
dd <- zoo(180/pi*atan2(coredata(uv)[,1],coredata(uv)[,2]),order.by=index(uv))
plot(annual(dd,FUN='mean'),lwd=2,ylim=c(-180,180),
     main='Annual mean wind direction (degrees)')
axis(4,at=c(-180,-90,0,90,180),labels=c('S','E','N','W','S'))
for (j in c(-180,-90,0,90,180)) lines(range(index(dd)),rep(j,2),lty=3)
```

The change in 2006-2010 is also visible in the wind direction. This could be an event caused by a change in circulation, but it cluld also be due to changes in instrumentation, data used in the gridding or observational practices. It is possible to examine whether there were changes in the criculation by using ESD to compare the windspeeds with SLP from reanalyses.

@@

```{r zonal-wind-check1, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Trend in upper wind direction
uam <- annual(subset(uv,is=2),nmin=300)
ds.u1 <- DS(uam,eof.slp,eofs=1:20)
plot(ds.u1)
```

The results suggest a mismatch between the reanalysis and the EOBS data, and the change after 2005 is not corroberated by other sources. We can therefore say that this feature is suspect.

```{r zonal-wind-check2, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Trend in upper wind direction
ds.u2 <- DS(subset(uam,it=c(1950,2005)),eof.slp,eofs=1:20)
plot(ds.u2)
```

When the test is repeated for the shorter period excluding data after 2005, there is a reasonably good match between the EOBS data and the reanalysis. The trend analysis can be repeated with this event excluded

```{r wind-speed-trend-2, warning=FALSE, message=FALSE}
## Trend in upper wind speed
plot(subset(annual(ff,"quantile",probs=0.98),it=c(1950,2005)),lwd=2,
     main='Wind speed: annual 98-percentile (m/s)')
lines(trend(subset(annual(ff,"quantile",probs=0.98),it=c(1950,2005))),lty=2)
grid()
```

The exclusion of the suspicious event also affects the long-term trend estiamte - now it's zero with a slight reduction.

@@

@@

>>>>

**Summary**

Synoptic storms (low-pressure systems) represent an important factor that may affect flooding and innuation in the Torbay region, as they are both associated with precipitation, wind and storm surges. Their influence depends on their intensity as well as the amount of rain that they bring. The storm statistics exhibits a seasonal dependency, with more frequent and intense storms during autumn and winter. The number of storms and intensity are sensitive to sea surface temperatures and large-scale atmospheric conditions such as the North Atlantic Oscillation and the location of the upper air jet. The severity of storm surges that they cause is increased with a sea-level rise.

Other phenomena that may cause local flooding can be connected to convective events and cloud bursts. Observations so far have suggested that the amount of precipitation from such local phenomena has been secondary to synoptic storms for places like Torbay.

@@

# Definition of functions

@@

@@

# The data analysis

## Precipitation

First, we examine historical observed precipitation to get a picture of what has happened until now. Global warming has taken place for a few decades (especially since the 1980s), and we ought to see some trends already as a concequence of an increased greenhouse effect. Past changes will give some indication of how sensitive the local consequences are to the large-scale changes. The following lines load the data and weed out stations with lots of missing values

@@

Need to fill in missing values to proceed with the analysis, but since there are too many missing values, we need to get some more data from the region. This is ok since we are working with annual aggregated values, which may imply larer spatial scales. In any case, these are used to assist the interpolation for the MIDAS sations.

@@

supplement ECA\&D to the MIDAS data:

@@

The five stations with the highest mean wet-day precipitation suggest a decline in the intensity between 1960-2010, but most of the stations with smaller means suggest an increase over the same period.

```{r}
plot(fw,new=FALSE); grid()
for (i in 1:dim(mu)[2]) lines(trend(fw[,i]))
```

The number of rainy days appears to be fairly stable with a frequency of around 0.35, or 120-130 days per year (it may rain for only a fraction of the day).

```{r probabilities, warning=FALSE, message=FALSE}
Pr.gt.30 <- PrexpPr(mu,fw,x0=30)
plot(Pr.gt.30,new=FALSE); grid()
for (i in 1:dim(mu)[2]) lines(trend(Pr.gt.30[,i]))
```

@@

Another attempt to search for systematic influences on the precipitation statistics ($\mu$) is to examine the mean seasonal cycle. This type of approach may give som indication of how sensitive precipitation is to changing conditions.

@@

```{r precipitation-intensity, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the wet-day mean precipitation
plot(MU,new=FALSE)
grid()
```

There is some indication of a mean seasonal cycle in $\mu$, with minimum values in late winter spring and higher values in late summer and autumn. This suggest that the intensity is driven more by precipitation associated with autumn storms (low-pressure systems) than summer-time convective events.

```{r wet-day frequency, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the wet-day frequency
plot(FW,new=FALSE)
grid()
```

There is a clear seasonal cycle in the wet-day occurrence - more rainy days in the autumn. This too may be connected to storms. The minimum tends to be in June-August when convective events are expected to be more dominant.

```{r sesonal-precipitation, warning=FALSE, message=FALSE}
## Show the mean seasonal cycle in the total precipitation
MP <- 30*MP
attr(MP,'unit') <- 'mm/month'
plot(MP,new=FALSE)
grid()
```

@@

```{r winter-rain, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Estimate the PCA for the annual cycle to identify the general features:
pca.MU <- PCA(MU,n=10)
plot(pca.MU,new=FALSE)
```

Most stations have the same features when it comes to a mean seasonal cycle in $\mu$, with the leading PCA accounting for 80% of the variance. There is some variations in the amplitude from place to place, and the mean precipitation is influenced by geography and the local environment.

@@

Empirical-statistical downscaling of the wet-day mean precipitation (intensity) suggests a slight increase for the future, although the analysis is very simple and far from perfect. The method is described in Benestatet al. (2017) <http://www.nat-hazards-earth-syst-sci.net/17/993/2017/>. This method was appropriate for convective events, but not suitable for precipitation associated with synoptic storms/low pressure systems (see below for storms).

## EOBS gridded daily precipitation

@@

```{r predictand, warning=FALSE, message=FALSE}
pr.eobs <- retrieve('data.ECAD/rr_0.25deg_reg.nc',lon=c(-6,-1),lat=c(50,51.5))
mu <- annual(pr.eobs,FUN='wetmean')
fw <- annual(pr.eobs,FUN='wetfreq')
```

@@

```{r trend, warning=FALSE, message=FALSE}
## The area mean wet-day mean precipitation
plot(aggregate.area(mu,FUN='mean'),new=FALSE)
## The area mean wet-day frequency
plot(aggregate.area(fw,FUN='mean'),new=FALSE)
## The early part of the record looks suspect:
mu <- subset(mu,it=c(1958,2015))
fw <- subset(fw,it=c(1958,2015))
## Show maps of trends for the wet-day mean precipitation
map(mu,FUN='trend',
    colbar=list(breaks=seq(-0.12,0.12,by=0.02),pal='t2m',rev=TRUE),new=FALSE)
## Map of trends in wet-day frequency:
map(fw,FUN='trend',
    colbar=list(breaks=seq(-0.01,0.01,by=0.001),pal='t2m',rev=TRUE),new=FALSE)
```

@@

```{r EOFs-in-precip-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The area mean wet-day mean precipitation
eof.mu <- EOF(mu)
## The area mean wet-day frequency
plot(eof.mu,new=FALSE)

@@

```{r EOFs-in-precip-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Show maps of trends for the wet-day mean precipitation
eof.fw <- EOF(fw)
## Map of trends in wet-day frequency:
plot(eof.fw,new=FALSE)
```

@@

The predictor describes larg-scale conditions on which the local precipitation is expected to depend. The occurrence (frequency) of precipitation depends om the atmospheric circulation pattern (captured by mean sea-level pressure - SLP - patterns). The intensity is assumed to be influenced by the atmospheric humidity, which is influenced by the rate of evaporation and hence maritime temperatures.

@@

```{r DS-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The wet-day frequency and SLP
ds.fw <- DS(subset(eof.fw,ip=1:4),eof.slp,eofs=1:5)
plot(ds.fw,new=FALSE)
```

```{r DS-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
ds.mu <- DS(subset(eof.mu,ip=1:4),eof.es,eofs=1:7)
plot(ds.mu,new=FALSE)
```

@@

```{r residual-for-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The residual of the wet-day precipitation
z.mu <- as.field(ds.mu)
#res.mu <- as.residual(z.mu)
res.mu <- mu - z.mu
res.mu <- attrcp(mu,res.mu)
class(res.mu) <- class(mu)
eof.res.mu <- EOF(res.mu)
plot(eof.res.mu,new=FALSE)
```

@@

For comparison the original data look like this:

```{r EOF-for-mu, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The original of the wet-day precipitation
plot(EOF(mu),new=FALSE)
```

```{r residual-for-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Repeat with the wet-day frequency for the benefit of comparison
z.fw <- as.field(ds.fw)
#res.fw <- as.residual(ds.fw)
## Another way to estimate the residuals:
res.fw <- fw - z.fw
res.fw <- attrcp(fw,res.fw)
class(res.fw) <- class(fw)
eof.res.fw <- EOF(res.fw)
plot(eof.res.fw,new=FALSE)
```

@@

For comparison the original data look like this:

```{r EOF-for-fw, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## The original of the wet-day precipitation
plot(EOF(fw),new=FALSE)
```

@@

Coastal innundation is expected to worsen with rising sea-levels, given similar types of mid-latitude storms as at present. To explore this type of effect, we can analyse the sea-level trends and connection to global temperatures to assess future outcome. First the stations from the GLOSS dataset (global monthly data):

@@

```{r}
plot(gloss,new=FALSE)
grid()
```

@@

```{r}
plot(newlyn,new=FALSE)
grid()
```

The curve suggests a longterm increase since the early 20th century, but the data stops before 2000.

@@

```{r sonel, warning=FALSE, message=FALSE}
sonel <- aggregate(station.sonel(),year,FUN='max',na.rm=TRUE)
plot(sonel,new=FALSE)
grid()
```

@@

```{r combine-sea-level, warning=FALSE, message=FALSE}
index(newlyn) <- year(newlyn)
index(sonel) <- year(sonel)
index(gloss) <- year(gloss)
Z <- combine.stations(newlyn,sonel,gloss)
nv <- apply(coredata(Z),2,FUN='nv')
Z <- subset(Z,is=(nv > 50))
Z <- subset(Z,it=c(1966,2015))
coredata(Z)[!is.finite(Z)] <- NA
coredata(Z) <- t(t(coredata(Z)) - colMeans(coredata(Z),na.rm=TRUE))
plot(Z,new=FALSE); grid()
```

There is a general increasing trend for the majority of the tidal stations, but the picture is clutterend by missing-data gaps, dupious outliers, and intermittent data availability.

@@

```{r sl-trend-map, warning=FALSE, message=FALSE, fig.height=8}
zc <- coredata(Z)
zc[is.infinite(zc)] <- NA
zc -> coredata(Z)
map(Z,pch=19,FUN='trend',cex=-2,add.text=TRUE,new=FALSE)
```

A trend analysis yields inconsistent results because missing-data gaps, of the inclusion of filled-in data by interpolation, and variable data quality. The map indicates different trends in the British Channel and the near-by parts of the Bay of Biscay.

@@

The PCA suggests stronger amplitude in the interannual variations in the sea-level near the Bay of Biscay, and less variation in the British Channel.

@@

Wind affects wave heights and can cause flooding and innundation through storm surges. Wind measurements are often unavailable, but it is possible to reconstruct wind informatnion from barometric pressure - sea-level pressure (SLP).

@@

```{r slp-trend-map, warning=FALSE, message=FALSE}
slp.eobs <- retrieve('data.ECAD/pp_0.25deg_reg.nc',lon=c(-6,0),lat=c(48,52))
slp.eobs <- 10000*slp.eobs ## The data have some funny scaling factor
attr(slp.eobs,'unit') <- 'Pa'
map(slp.eobs,FUN='trend',new=FALSE)
```

The trend analysis indicates non-zero changes near the British Channel, although these may be due to natural variability.

@@

```{r trend-map, warning=FALSE, message=FALSE}
## Fudge it - pick some windstations, but do not use their data. Reset the
## coordinates to select a triangle of SLP from the EOBS data.
ss <- select.station(param='slp',src='ecad',cntr=c('France','UNITED KINGDOM'))
fx <- station(ss)
triangle <- subset(fx,is=1:3)
## The stations are not situated quite where there is valid slp data in the EOBS dataset
## It's only their coordinates that is used for interpolating SLP for three stations
## Using station objects as is-object makes the results a station object
attr(triangle,'longitude') <- c(-4,-2,-1.25)
attr(triangle,'latitude') <- c(50.50,51,49)
pp.eobs <- regrid(slp.eobs,is=triangle)
## Use triagulation for stations to estimate the wind
uv <- geostrophicwind(pp.eobs)
plot(annual(uv,nmin=300),main='zonal and meridional wind components',
     sub='Estimated from triangulation of mean sea-level pressure from EOBS',zoom=7,new=FALSE)
grid()
```

THe level of the wind components seam fairly stable between 1950 and 2050, albeit with a number of peaks.

This event also is seen in the upper tail of the wind speed distribution. There are four features concerning these wind speed estimates: a downward trend from moderate high wind speeds in the 1950s, two years with a large drop in values in the 1970s, a number of years with low values around year 2010, and a slight decreasing trend in the upper wind speeds over the record.

```{r wind-speed-trend, warning=FALSE, message=FALSE}
## Trend in upper wind speed
ff <- sqrt(uv[,1]^2 + uv[,2]^2)
plot(annual(ff,"quantile",probs=0.98),lwd=2,
     main='Wind speed: annual 98-percentile (m/s)',new=FALSE)
lines(trend(annual(ff,"quantile",probs=0.98)),lty=2)
grid()
trend.coef(annual(ff,"quantile",probs=0.98))
trend.err(annual(ff,"quantile",probs=0.98))
```

@@

```{r wind-direction-trend, warning=FALSE, message=FALSE}
## Trend in upper wind direction
dd <- zoo(180/pi*atan2(coredata(uv)[,1],coredata(uv)[,2]),order.by=index(uv))
plot(annual(dd,FUN='mean'),lwd=2,ylim=c(-180,180),
     main='Annual mean wind direction (degrees)',new=FALSE)
axis(4,at=c(-180,-90,0,90,180),labels=c('S','E','N','W','S'))
for (j in c(-180,-90,0,90,180)) lines(range(index(dd)),rep(j,2),lty=3)
```

It is possible to examine whether there were changes in the criculation by using ESD to compare the windspeeds with SLP from reanalyses.

@@

```{r zonal-wind-check1, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
## Trend in upper wind direction
uam <- annual(subset(uv,is=2),nmin=300)
ds.u1 <- DS(uam,eof.slp,eofs=1:20)
plot(ds.u1,new=FALSE)
```

@@

The windspeed statistics has a distribution that is close to being Weibul, although with a slight bias.

@@

## Storms

Synoptic storms are associated with winds, storm surges, precipitation, and waves. The storms (low pressure systems) tend to appear over the North Atlantic ("cyclogenesis"") between the coast of Newfoundland and the British Isles, and travel eastward and die off once they hit land (Scandinavia). These storms are a result of wind shear and temperature differences, due to barolinic instability, and tend to cluset around the upper air jet stream. The storm statistics is also influenced by the North Atlantic Oscillation (NAO). Since the storms travel a fair distance over the North Atlantic, they are expected to be subject to air-sea heat and moisture exchange before hitting the UK. The intensity of the storms may potentially be affected by affected by sea surface temperatures through air-sea interactions - higher temperatures favour more evaporation and more "fuel" for the storms.

A plot of the number of trajectories over time suggests a weak declining trend over 1980-2015 in the number of storms sweeping over the Torbay region.

```{r}
plot(cyclones.uk,new=FALSE)
grid()
```

This curve does not indicate how the mean intensity of the storms has varied over the same period.

```{r}
cdat <- as.character(cyclones.uk[,2])
press <- annual(zoo(apply(cyclones.uk[,is.element(colnames(cyclones.uk),'pcent')],1,min),
             order.by=as.Date(paste(substr(cdat,1,4),'-',substr(cdat,5,6),'-',
                                    substr(cdat,7,8),':',substr(cdat,9,10),':00',sep=''))),'min')
plot(press,main='Annual minimum central storm sea-level pressure',ylab='pressure (hPa)')
grid()
```

The central pressure is more closely connected to the wind speeds in the storm systems, but the connection between amount of precipitation caused by such storms is less clear. The precipitation intensity is expected to be influenced by the air moisture and the rate of evaporation over the ocean where the storms cross.
